### 2024-05-17 Manuscript/main.tex
Update main.tex

In the last two decades, there has been a progressive shift in ecology and evolution from  taxonomic-oriented approaches, with species as the focal point of studies, to functional approaches that place species-specific characteristics (traits) at the centre of analyses \citep{mammola2021concepts, palacio2022protocol}).
Although numerous definitions exist \citep{dawson2021traits}, we here consider a trait to be any characteristic - e.g. , morphological, anatomical, ecological, physiological, behavioural, phenological—measured on individual organisms at any level, from genes to whole organisms.
A trait-based approach has two advantages for answering core questions in ecology and evolution.
First, it allows a deeper understanding of the mechanisms generating biodiversity patterns, by putting organisms’ traits at the centre of natural selection rather than organisms themselves.
Second, it allows comparisons across subdisciplines in biology, while facilitating the conceptualization of general principles broadly valid in space (e.g. unrelated species pools) and time (e.g. anatomical traits are comparable between palaeontology and ecology, which is not always the case with species; \citealt{luza2023going}).
Already used routinely in palaeontology \cite{raup1961geometry, gould1991disparity, foote1995morphological, guillerme2020disparities}, this new trait-focused ecology and evolution is unlocking the possibility to answer a broad range of questions in disciplines as diverse as community ecology \cite{mcgill2006rebuilding}, biogeography \cite{violle2014emergence}, conservation biology \cite{chichorro2022trait}, micro- \cite{chapin1993evolution}, macro-evolution \cite{wiens2023trait} %TG: check to find another cite
 and applied fields (e.g. agronomy \citealt{martin2015plant}).
This is because trait-based approaches closely align with the general analytical framework proposed by \cite{anand1994pattern} to answer three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism) (Box 1).

@@

\section*{Box}
\cite{anand1994pattern} conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism)
This framework can be effectively applied to trait-based analyses:

@@

Note that the distinction between what, how and why is not categorical in its nature and is often nuanced, with patterns and process or process and mechanisms sometimes being used interchangeably to describe the same questions

@@

While the first part of this framework (``what?'') has been studied extensively in the last decade (@CITECITECITE), %TG: TODO
 this was often undertaken outside of the ``what, how and why'' framework.
Here we argue that the pattern description, although the basis of any future study, often follows some framework without a proper evaluation of its adequacy to answer the how and why.
The choice of the tool or metric used to describe the pattern is crucial for allowing us to understand the process and the mechanism.
Using an inappropriate metric for describing a pattern can lead to biased conclusions.
For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - ``why?'') on an island compared to another island with only one of the species (the process - ``how?'').
In such a scenario, measuring the occupied area of each population on each island (a pattern - ``what?'') will not be the most appropriate way to understand the potential competition between these populations.
In this simplistic example, the functional overlap between populations (``what and how'') might be more appropriate.
See for example \citealt{carvalho2020decomposing} for a discussion of how Darwin’s finches share resources depending on the existence or not of competition.

@@

We then applied five different stressors to the non-stressed trait space to remove either 20\%, 40\%, 60\% or 80\% of the data (resulting in trait spaces with, respectively, 160, 120, 80 and 40 elements).
In removing the data, we applied these five specific algorithms (Figure \ref{Fig:simulations}; Table \ref{Tab:mechanisms}; all algorithms, except ``Evenness'', were previously described in \cite{guillerme2020shifting}):

@@

\item Increasing density: by removing the required amount of pairs of points with a pair distance of at least $D$ (i.e. removing the n pairs of points that are at least $D$ distance away from each other).
This approximates our \textbf{facilitation mechanism} (i.e. the points left are only ones that are close to at least another point in space).

@@

\item Increasing evenness: by resampling the required amount of data but with skewed resampling probabilities per bandwidth (i.e. an estimated number of categories summarising the distribution, like in a histogram).
Each bandwidth has an observed probability of sampling of n (proportional to the number of elements in that bandwidth) and is skewed to a probability of sampling of $n \times m^{p}$ where $m>0$ for the low values of $n$ and $m<0$ for high values of $n$ (respectively low and high density bandwidths) for flattening the curve (and the opposite for steepening it) and p is a factor increasing the flattening/steepening (here we used $p=3$).
This approximates our \textbf{competition mechanism}.
\end{itemize}

For the position, size and density algorithms, we estimated the parameters $\rho$ and $D$ recursively with the \texttt{dispRity} package to obtain the required amount of data to be removed (\texttt{dispRity::reduce.space}, \citealt{guillerme2018disprity,guillerme2020shifting}).
These steps resulted in 4000 simulated trait spaces (200 simulated spaces $\times$ 5 stressors $\times$ 4 amounts of data removed).
See Table \ref{Tab:mechanisms} for a biological description of these stressors and Figure \ref{Fig:simulations} for a visual description of them.

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.5\textwidth]{Figures/simulation_protocol.pdf}
\caption{\scriptsize{Illustration of the simulation protocol used in this paper.
A) we simulated a pure birth-tree with a trait evolving under a Brownian Motion process until reaching 200 tips resulting in a one dimensional trait space or two dimensional space.
We then applied different stressors to the resulting trait space using the following:
B) \textbf{null mechanism}: randomly removing half of the species resulting in two groups of 100 tips (blue and orange - the brown colour in the histogram represents the overlap between both groups);
C) \textbf{equalising mechanism}: removing species on the edge of the distribution;
D) \textbf{facilitation mechanism}: removing species to reduce the distance between pairs of species (increasing local density; decreasing evenness);
E) \textbf{filtering mechanism}: removing species on one extreme of the distribution;
F) \textbf{competition mechanism}: proportionally removing species on the centre of the distribution (``flattening the curve'').}}
\label{Fig:simulations}
\end{figure}
\bigskip

@@

We structured our simulations based on three aspects of diversity commonly captured by functional diversity metrics \citep{mammola2021concepts}:
1) Richness, encompassing metrics reflecting the sum of difference among observations; (4 metrics, equivalent to size metrics in \citealt{guillerme2020shifting});
2) Divergence, encompassing metrics reflecting the average difference among observations (4 metrics); and
3) Regularity, encompassing metrics reflecting how regular the differences among observations are (4 metrics, equivalent to density metrics in \citealt{guillerme2020shifting}).
Note that we focused here on three packages fully devoted to functional diversity analyses (\texttt{BAT} \citealt{cardoso2015bat} \texttt{FD} \citealt{Laliberte2010FD} and \texttt{TPD} \citealt{carmona2019trait}) in the \texttt{R} statistical environments (see \citealt{mammola2021concepts}).

@@

\textbf{Aspect} & \textbf{Method} & \textbf{Name} & \textbf{Description} & \textbf{\texttt{R} function} & \textbf{Reference}\\
\hline
 & Trait Probability Density & Functional evenness & @@@NEEDS DESCRIPTION & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional evenness &  Overlap between a hypervolume and an imaginary hypervolume where traits are evenly distributed within their possible range & \texttt{BAT::kernel.evenness} & \cite{mammola2020functional}\\
 & \cr 
 & Dendrogram & Functional evenness & Regularity of abundances and distances between observations in the tree. & \texttt{BAT::evenness} & \cite{cardoso2015bat,cardoso2024calculating}\\
 & \cr 
 & Distances & Functional evenness & Regularity of abundances and distances along the minimum spanning tree linking all observations & \texttt{FD::dbFD()\$FEve} & \cite{Laliberte2010FD,villeger2008new}\\
\hline
Divergence & Trait Probability Density & Divergence & Overlap between the TPD function of a single observation and the TPD function of the whole set of observations & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional dispersion  & Average distance between an observation and a sample of random points within and the probabilistic hypervolume & \texttt{BAT::kernel.dispersion} & \cite{mammola2020functional}\\
 & \cr 
 & Dendrogram & Functional dispersion & Average dissimilarity between any two observations in the dendrogram & \texttt{BAT::dispersion} & \cite{cardoso2015bat}\\
 & \cr  
 & Distances & Rao's quadratic entropy & @@@NEEDS DESCRIPTION & \texttt{FD::dbFD()\$RaoQ} & \cite{botta2005rao}\\
\hline
Richness & Trait Probability Density & Functional richness & Sum of cells where trait probability density $> 0$ & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional richness & Size (volume) of the hypervolume & \texttt{BAT::kernel.alpha} & \cite{mammola2020functional}\\
 & \cr
 & Convex Hull & Functional richness & Volume of the convex hull & \texttt{BAT::hull.alpha} & \cite{cornwell2006trait}\\
 & \cr
 & Dendrogram & Alpha diversity & Total branch length of the functional dendrogram & \texttt{BAT::alpha} & \cite{petchey2002functional,cardoso2015bat}\\
\hline

@@

To ease interpretation of the results, we centred the results for the 14 metrics for the 4 non null stressors by the results for the null stressor and then scaled them.
Effectively, for each metric and each single simulation, we subtracted the results under the null stressor to the ones under the other stressors and divided the resulting centred metrics by the maximum value for each metric between each simulation.
This resulted in each metric being scaled between –1 and 1 where 0 corresponds to no change compared to the null and 1 and –1 the maximum positive and negative deviations relative to the null.
In parallel, we also compared the distribution of the 200 scores for each metric with each reduction level for each non-null stressor to the null stressor (with no scaling or centering) using pairwise t-tests.
The results of these tests are displayed in table @@@ %TG: TODO: add table
 and in figure \ref{Fig:simulation_results} (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).

@@

We applied all metrics to an empirical dataset of bird extinctions on six Hawaiian Islands (Hawaii, Kauai, Lanai, Maui, Molokai, Oahu), an archipelago which has suffered large numbers of anthropogenic extinctions due to a range of extinction drivers \citep{Walther2022}.
A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at 1500 CE) and extant species list of the native community for each island was taken from \cite{matthews2023global} along with trait values for the extinct species (Asio flammeus was removed from the historic and extant species lists as its colonisation status is uncertain).
We also added two marine species to the prehistoric species list; we compiled missing trait data for these two extinct species from @ferrans data
We extracted traits for extant species from the AVONET database \citep{tobias2022avonet}
We used 9 morphological traits associated with dietary/foraging preference and dispersal ability \citep{pigot2020macroevolutionary,sheard2020ecological} to represent the functional diversity of the communities: mass; beak length (culmen); beak length (nares); beak width; beak depth; tarsus length; wing length; kipp’s distance (length from the first secondary feather to the tip of the longest primary feather; \citealt{tobias2022avonet}); and tail length
Traits were log transformed and standardised to a mean of 0 and standard deviation of 1, prior to analyses.

@@

The ability of different metrics to capture the different patterns (and thus approximate the processes) was highly variable.
It ranged from metrics capturing no clear pattern (e.g. the Functional richness based on the hypervolume method) to metrics clearly capturing one specific pattern (e.g. Rao’s quadratic entropy based on pairwise distances capturing clear changes in the equalising mechanism but less clear changes for the other mechanisms).
Some metrics even captured an opposite pattern to what was expected: for all metrics, we expected a relative decrease in trait space occupancy following different levels of species removals but some metrics captured an \textit{increase} of trait space occupancy (e.g.
Functional richness using the convex hull method for equalizing) or non-linear ones (e.g. Functional richness using the probability density method for facilitation).
This raises an interesting point: the naive expectation that removing 80 to 20\% of the data would result in a reduction of trait space occupancy metric score (negative values) is often true but by no means always true.
Some cases resulted in an increase in metric score (positive values)!

@@

For the empirical data, we see again a range of different possible interpretations depending on the metric describing the pattern.
For example, all richness metrics (green; Figure 3) pick up a similar relative increase (i.e., relative to the null stressor) in richness for both the historic and extant datasets.
However, the results are occasionally contrasting between the historical and extant datasets, with a respective decrease and increase with some divergence metrics (e.g.
Function dispersion based on dendrograms or hypervolumes).
Finally, some metrics pick up a decrease for both datasets (e.g.
Functional evenness based on hypervolumes).
Note that this apparent global increase in most metric values may appear counterintuitive given that 46\% and 69\% of species went extinct over the two time periods.
This is due to the scaling of the metrics compared to random equivalent extinctions.
That is, Figure 3 is not displaying absolute changes in trait space occupancy but rather changes in trait space occupancy relative to a crude neutral model where all species are equally likely to go extinct.
%TG: Move below to discussion?
However, this is still an intriguing finding, as it indicates that, at least for certain metrics, the loss of Hawaiian bird FD is less than expected than if extinctions were random with respect to traits.
This contrasts with studies that have shown that species with certain traits (e.g., larger species) are more likely to have gone extinct and thus that FD loss is larger than through random extinctions (e.g., \citealt{sayol2021loss,Matthews2022}).
This could partly be due to the proportion of passerine extinctions in Hawaii.
Although most global bird species are (or were, if accounting for anthropogenic extinctions) passerines ($\sim$58\%), the majority of global known anthropogenic extinctions ($\sim$75\%) have been non-passerines, which tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in FD).
However, in Hawaii, where the same proportion of known species were passerines ($\sim$58\%), the opposite is true: the majority of extinctions ($\sim$60\%) have been of passerine species.
It is important to note that we are not implying that there existed little morphological variation within extinct Hawaiian passerines – indeed, many of these extinctions involved the honeycreepers, a famous adaptive radiation involving substantial evolution of beak morphology \citep{Walther2022} – but simply that this morphological variation is less than that observed across all birds.
But these results – given that the proportion of known Hawaiian species that are / were passerines roughly matches the proportion of extinct species that were passerines –  also indicate that these extinct species were more functionally similar than expected, thus relatively \textit{increasing} trait diversity (i.e.
relative to random extinctions).
% @@@**add sentence about the large number of rail extinctions**.
In addition, it is worth stressing that these analyses included marine species (which constitutes less than 4\% of  known extinctions in Hawaii, while making up 11\% of the prehistoric assemblage) and different results may have been found if focusing exclusively on terrestrial birds.

More broadly, as can be seen from the Hawaiian case study, patterns are likely to be much less clear with empirical data due to multiple stressors acting simultaneously.
%TG: Move above to discussion?
For instance, extensive hunting is likely to present equalising and/or filtering pressure due to the selective hunting of larger bodied species.
In contrast, avian malaria, which is known to be an important extinction pressure of native Hawaiian birds \cite{samuel2011dynamics}, could show signals of a null mechanism since susceptibility to the disease is not trait-dependent (at least not in relation to the traits we have considered here), but rather genetic susceptibility (it mainly affects passerines, but there is considerable variation in susceptibility within passerines) and distributional range \citep{jessup2023wildlife}.

@@

Furthermore, we only presented results based on a relatively small number of dimensions (up to 8), while it is not uncommon to have a much greater number of dimensions, especially in paleontology (@@@ examples).
In a higher number of dimensions (usually >10 but this is highly variable depending on the system under study), metrics results are harder to scale linearly.
This sometimes makes metric interpretation more chaotic (@curse\_of\_dimensionality) in the sense that there is no direct relationship between the number of observations and dimensions that can be used to extrapolate results in higher dimensional datasets.
In fact, the relationship between observations has an important effect on how our results scale with higher dimensionality (e.g.
the volume of an evenly occupied space scales differently in high dimensions compared to the same space but with normally distributed observations).
Therefore we advise workers to test assumptions specifically about their trait space and ``play'' around with it before choosing their functional diversity/disparity framework and metric of interest.

@@

Following our results and these thoughts, we would like to encourage workers in ecology and evolution to test whether their metrics are capable of capturing a pattern of interest before discussing the process and/or mechanism inferred by it.
We encourage workers to test their metrics using simplified example datasets to see if it has the ability to capture at least crude changes as exemplified in this work (e.g. using simulations \citealt{guillerme2024treats} or the \texttt{moms} interactive package) \citealt{guillerme2020shifting}.
Importantly, it should always be borne in mind that empirical mechanisms ought of course to be more complex in real world scenarios.
For example, evolutionary mechanisms can vary through time or per clades (and so patterns are often the results of multiple processes); or ecological mechanisms are often intertwined and work together to generate a pattern (e.g.
facilitation + filtering; @ref), %TG: TODOL find a reference
 or counteract each others by operating on the trait space in opposite directions (e.g., competition + filtering; \citealt{mammola2024functional})

\section{Conclusion}
Different ecological and evolutionary processes or mechanisms do not always lead to different patterns.
Different observed patterns can originate from the same processes and mechanisms.
Our results based on simulations and empirical data suggest that the same data can be interpreted differently depending on the choice of trait space occupancy metric (disparity, dissimilarity or functional diversity metrics).
Different metrics are designed to capture different aspects \citep{guillerme2020shifting,mammola2021concepts} but also perform better or worse at their designed task depending on the mechanism or process of interest.
In other words, it is important to choose a disparity/dissimilarity/functional diversity metric depending on the question and the data at hand rather than as a default option or based on a previous inspiring publication.

@@

\section{Acknowledgements}
Thanks to Andrew Beckerman, Natalie Cooper, Alain Danet, Thomas Johnson and Gavin Thomas for comments on early versions of this manuscripts.\\
Cite grants

>>>>

In the last two decades, there has been a progressive shift in ecology and evolution from  taxonomic-oriented approaches, with species as the focal point of studies, to functional approaches that place species-specific characteristics (traits) at the centre of analyses \citep{mammola2021concepts, palacio2022protocol}.
Although numerous definitions exist \citep{dawson2021traits}, we here consider a trait to be any observable characteristic (e.g. morphological, anatomical, ecological, physiological, behavioural, phenological, etc.) measured on individual organisms at any level, from genes to whole organisms.
A trait-based approach has two advantages for answering core questions in ecology and evolution.
First, it allows a deeper understanding of the mechanisms generating biodiversity patterns, by putting organisms’ traits at the centre of natural selection rather than organisms themselves.
Second, it allows comparisons across subdisciplines in biology, while facilitating the conceptualization of general principles broadly valid in space (e.g. unrelated species pools) and time (e.g. anatomical traits are comparable between palaeontology and ecology, which is not always the case with species; \citealt{luza2023going}).
Already used routinely in palaeontology \citep{raup1961geometry, gould1991disparity, foote1995morphological, guillerme2020disparities}, this new trait-focused ecology and evolution is unlocking the possibility to answer a broad range of questions in disciplines as diverse as community ecology \citep{mcgill2006rebuilding}, biogeography \citep{violle2014emergence}, conservation biology \citep{chichorro2022trait}, micro- \citep{chapin1993evolution}, macro-evolution \citep{guillerme2023innovation} and applied fields (e.g. agronomy \citealt{martin2015plant}).
This is because trait-based approaches closely align with the general analytical framework proposed by \cite{anand1994pattern} to answer three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism) (see Box 1).

@@

\section*{Box 1}
\cite{anand1994pattern} conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism).
This framework can be effectively applied to trait-based analyses:

@@

Note that the distinction between what, how and why is not categorical in its nature and is often nuanced, with patterns and process or process and mechanisms sometimes being used interchangeably to describe the same questions.

@@

While the first part of this framework (``what?'') has been studied extensively in the last decade this was often undertaken outside of the ``what, how and why'' framework.
Here we argue that the pattern description, although the basis of any future study, often follows some framework without a proper evaluation of its adequacy to answer the how and why.
The choice of the tool or metric used to describe the pattern is crucial for allowing us to understand the process and the mechanism.
Using an inappropriate metric for describing a pattern can lead to biased conclusions.
For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - ``why?'') on two islands, one home two both populations and another one home to only one of the population (the process - ``how?'').
In such a scenario, measuring the occupied area (i.e. $km^{2}$) of each population on each island (a pattern - ``what?'') will not be the most appropriate way to understand the potential competition between these populations.
In this simplistic example, the functional overlap between populations (``what and how'') might be more appropriate.
See for example \citealt{carvalho2020decomposing} for a discussion of how Darwin’s finches share resources depending on the existence or not of competition.

@@

We then applied four different stressors to the non-stressed trait space to remove either 20\%, 40\%, 60\% or 80\% of the data (resulting in trait spaces with, respectively, 160, 120, 80 and 40 elements).
In removing the data, we applied these five specific algorithms (Figure \ref{Fig:simulations}; Table \ref{Tab:mechanisms}; all algorithms, except ``Evenness'', were previously described in \cite{guillerme2020shifting}):

@@

\item Increasing density: by removing the required amount of pairs of points with a pair distance of at least $D$ (i.e. removing the $n$ pairs of points that are at least $D$ distance away from each other).
This approximates our \textbf{facilitation mechanism} (i.e. the points left are only ones that are close to at least another point in space).

@@

\item Increasing evenness: by resampling the required amount of data but with skewed resampling probabilities per bandwidth (i.e. an estimated number of categories summarising the distribution, like in a histogram).
Each bandwidth has an observed probability of sampling of $n$ (proportional to the number of elements in that bandwidth) and is skewed to a probability of sampling of $n \times m^{p}$ where $m>0$ for the low values of $n$ and $m<0$ for high values of $n$ (respectively low and high density bandwidths) for flattening the curve (and the opposite for steepening it) and $p$ is a factor increasing the flattening/steepening (here we used $p=3$).
This approximates our \textbf{competition mechanism}.
\end{itemize}

For the position, size and density algorithms, we estimated the parameters $\rho$ and $D$ recursively with the \texttt{dispRity} package to obtain the required amount of data to be removed (\texttt{dispRity::reduce.space}, \citealt{guillerme2018disprity,guillerme2020shifting}).
These steps resulted in 3420 simulated trait spaces (171 simulated spaces $\times$ 5 stressors $\times$ 4 amounts of data removed).
We used 171 replicates per simulation because that was empirically the smallest number of simulation required to reach a variance between simulation lower than 1\% across all metrics (i.e. any additional simulation added less that 1\% extra variance).
See Table \ref{Tab:mechanisms} for a biological description of these stressors and Figure \ref{Fig:simulations} for a visual description of them.

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.5\textwidth]{Figures/simulation_protocol.pdf}
\caption{\scriptsize{Illustration of the simulation protocol used in this paper.
A) we simulated a pure birth-tree with a trait evolving under a Brownian Motion process until reaching 200 tips resulting a trait space of 2, 4, or 8 dimensions.
We then applied different stressors to the resulting trait space using the following:
B) \textbf{null mechanism}: randomly removing half of the species resulting in two groups of 100 tips (blue and orange - the brown colour in the histogram represents the overlap between both groups);
C) \textbf{equalising mechanism}: removing species on the edge of the distribution;
D) \textbf{facilitation mechanism}: removing species to reduce the distance between pairs of species (increasing local density; decreasing evenness);
E) \textbf{filtering mechanism}: removing species on one extreme of the distribution;
F) \textbf{competition mechanism}: proportionally removing species on the centre of the distribution (``flattening the curve'').}}
\label{Fig:simulations}
\end{figure}
\bigskip

@@

We structured our simulations based on three aspects of diversity commonly captured by functional diversity metrics \citep{mammola2021concepts}:
1) Richness, encompassing metrics reflecting the sum of differences among observations; (4 metrics, equivalent to size metrics in \citealt{guillerme2020shifting});
2) Divergence, encompassing metrics reflecting the average differences among observations (4 metrics); and
3) Regularity, encompassing metrics reflecting how regular the differences among observations are (4 metrics, equivalent to density metrics in \citealt{guillerme2020shifting}).
Note that we focused here on three packages fully devoted to functional diversity analyses (\texttt{BAT} \citealt{cardoso2015bat} \texttt{FD} \citealt{Laliberte2010FD} and \texttt{TPD} \citealt{carmona2019trait}) in the \texttt{R} statistical environments (see \citealt{mammola2021concepts}).

%  STOPPED HERE

@@

\textbf{Aspect} & \textbf{Method} & \textbf{Name} & \textbf{Description} & \textbf{\texttt{R} function} & \textbf{Reference}\\
\hline
 & Trait Probability Density & Functional evenness & An index of the spread of the trait distribution in an even spread of the distribution (e.g. the height of bins in a histogram) & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional evenness &  Overlap between a hypervolume and an imaginary hypervolume where traits are evenly distributed within their possible range & \texttt{BAT::kernel.evenness} & \cite{mammola2020functional}\\
 & \cr 
 & Dendrogram & Functional evenness & Regularity of abundances and distances between observations in the tree. & \texttt{BAT::evenness} & \cite{cardoso2015bat,cardoso2024calculating}\\
 & \cr 
 & Distances & Functional evenness & Regularity of abundances and distances along the minimum spanning tree linking all observations & \texttt{FD::dbFD()\$FEve} & \cite{Laliberte2010FD,villeger2008new}\\
\hline
Divergence & Trait Probability Density & Divergence & Overlap between the TPD function of a single observation and the TPD function of the whole set of observations & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional dispersion  & Average distance between an observation and a sample of random points within and the probabilistic hypervolume & \texttt{BAT::kernel.dispersion} & \cite{mammola2020functional}\\
 & \cr 
 & Dendrogram & Functional dispersion & Average dissimilarity between any two observations in the dendrogram & \texttt{BAT::dispersion} & \cite{cardoso2015bat}\\
 & \cr  
 & Distances & Rao's quadratic entropy & The mean Euclidean distance between every two random pairs of species & \texttt{FD::dbFD()\$RaoQ} & \cite{botta2005rao}\\
\hline
Richness & Trait Probability Density & Functional richness & Sum of cells where trait probability density $> 0$ & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional richness & Size (volume) of the hypervolume & \texttt{BAT::kernel.alpha} & \cite{mammola2020functional}\\
 & \cr
 & Convex Hull & Functional richness & Volume of the convex hull & \texttt{BAT::hull.alpha} & \cite{cornwell2006trait}\\
 & \cr
 & Dendrogram & Alpha diversity & Total branch length of the functional dendrogram & \texttt{BAT::alpha} & \cite{petchey2002functional,cardoso2015bat}\\
\hline

@@

To ease interpretation of the results, w e centred the results for the 14 metrics for the 4 non null stressors by the results for the null stressor and then scaled them.
Effectively, for each metric and each single simulation, we subtracted the results under the null stressor to the ones under the other stressors and divided the resulting centred metrics by the maximum value for each metric between each simulation.
This resulted in each metric being scaled between –1 and 1 where 0 corresponds to no change compared to the null and 1 and –1 the maximum positive and negative deviations relative to the null.
In parallel, we also compared the distribution of the 200 scores for each metric with each reduction level for each non-null stressor to the null stressor (with no scaling or centering) using pairwise t-tests.
The results of these tests are displayed in supplementary tables 1 to 3 and in figure \ref{Fig:simulation_results} (for the non-null stressor, grey distributions are not clearly distinguishable from the null stressor).

@@

We applied all metrics to an empirical dataset of bird extinctions on six Hawaiian Islands (Hawaii, Kauai, Lanai, Maui, Molokai, Oahu), an archipelago which has suffered large numbers of anthropogenic extinctions due to a range of extinction drivers \citep{Walther2022}.
A prehistoric species list (avifauna known to be present prior to human colonisation of the islands), historic species list (avifauna known to be present at 1500 CE) and extant species list of the native community for each island was taken from \cite{matthews2023global} along with trait values for the extinct species (Asio flammeus was removed from the historic and extant species lists as its colonisation status is uncertain).
We also added two marine species to the prehistoric species list; we compiled missing trait data for these two extinct species from \cite{sayol2021loss} data.
We extracted traits for extant species from the AVONET database \citep{tobias2022avonet}
We used 9 morphological traits associated with dietary/foraging preference and dispersal ability \citep{pigot2020macroevolutionary,sheard2020ecological} to represent the functional diversity of the communities: mass; beak length (culmen); beak length (nares); beak width; beak depth; tarsus length; wing length; kipp’s distance (length from the first secondary feather to the tip of the longest primary feather; \citealt{tobias2022avonet}); and tail length
Traits were log transformed and standardised to a mean of 0 and standard deviation of 1, prior to analyses.

@@

The ability of different metrics to capture the different patterns (and thus approximate the processes) was highly variable.
It ranged from metrics capturing no clear pattern (e.g. the Functional evenness based on the dendrogram method for competition) to metrics clearly capturing one specific pattern (e.g. Divergence based on probability density for the equalizing mechanism for a decrease in space occupancy).
Some metrics even captured an opposite pattern to what was expected: for all metrics, we expected a relative decrease in trait space occupancy following different levels of species removals but some metrics captured an \textit{increase} of trait space occupancy (e.g. the functional evenness based on hypervolume for and increase) or non-linear ones (e.g. Alpha diversity based on dendrogram for the equalizing mechanism).
This raises an interesting point: the naive expectation that removing 80 to 20\% of the data would result in a reduction of trait space occupancy metric score (negative values) is often true but by no means always true.
Some cases resulted in an increase in metric score relative to random removals (positive values in Figure \ref{Fig:simulation_results})!

@@

For the empirical data, we see again a range of different possible interpretations depending on the metric describing the pattern.
For example, all richness metrics (green; Figure 3) pick up a similar relative increase (i.e., relative to the null stressor) in richness for both the historic and extant datasets.
However, the results are occasionally contrasting between the historical and extant datasets, with a respective decrease and increase with some divergence metrics (e.g.
Function dispersion based on dendrograms or hypervolumes).
Finally, some metrics pick up a decrease for both datasets (e.g.
Functional evenness based on hypervolumes).
Note that this apparent global increase in most metric values may appear counterintuitive given that 46\% and 69\% of species went extinct over the two time periods.
This is due to the scaling of the metrics compared to random equivalent extinctions.
That is, Figure 3 is not displaying absolute changes in trait space occupancy but rather changes in trait space occupancy relative to a crude neutral model where all species are equally likely to go extinct.

More broadly, as can be seen from the Hawaiian case study, patterns are likely to be much less clear with empirical data due to multiple stressors acting simultaneously.
For instance, extensive hunting is likely to present equalising and/or filtering pressure due to the selective hunting of larger bodied species.
In contrast, avian malaria, which is known to be an important extinction pressure of native Hawaiian birds \cite{samuel2011dynamics}, could show signals of a null mechanism since susceptibility to the disease is not trait-dependent (at least not in relation to the traits we have considered here), but rather genetic susceptibility (it mainly affects passerines, but there is considerable variation in susceptibility within passerines) and distributional range \citep{jessup2023wildlife}.

@@

Furthermore, we only presented results based on a relatively small number of dimensions (up to 8), while it is not uncommon to have a much greater number of dimensions, especially in paleontology (e.g. in \citealt{van2023should}).
In a higher number of dimensions (usually >10 but this is highly variable depending on the system under study), metrics results are harder to scale linearly.
This sometimes makes metric interpretation more chaotic \cite{bellman1957dynamic} in the sense that there is no direct relationship between the number of observations and dimensions that can be used to extrapolate results in higher dimensional datasets.
In fact, the relationship between observations has an important effect on how our results scale with higher dimensionality (e.g. the volume of an evenly occupied space scales differently in high dimensions compared to the same space but with normally distributed observations).
Therefore we advise workers to test assumptions specifically about their trait space and ``play'' around with it before choosing their functional diversity/disparity framework and metric of interest.

@@

The empirical results are also intriguing, as it indicates that, at least for certain metrics, the loss of Hawaiian bird functional diversity is less than expected than if extinctions were random with respect to traits (Figure \ref{Fig:empirical_results}).
This contrasts with studies that have shown that species with certain traits (e.g., larger species) are more likely to have gone extinct and thus that functional diversity loss is larger than through random extinctions (e.g., \citealt{sayol2021loss,Matthews2022}).
This could partly be due to the proportion of passerine extinctions in Hawaii.
Although most global bird species are (or were, if accounting for anthropogenic extinctions) passerines ($\sim$58\%), the majority of global known anthropogenic extinctions ($\sim$75\%) have been non-passerines, which tend to be larger and possess more unique morphological traits (thus resulting in relatively large reductions in functional diversity).
However, in Hawaii, where the same proportion of known species were passerines ($\sim$58\%), the opposite is true: the majority of extinctions ($\sim$60\%) have been of passerine species.
It is important to note that we are not implying that there existed little morphological variation within extinct Hawaiian passerines - indeed, many of these extinctions involved the honeycreepers, a famous adaptive radiation involving substantial evolution of beak morphology \citep{Walther2022} - but simply that this morphological variation is less than that observed across all birds.
But these results, given that the proportion of known Hawaiian species that are / were passerines roughly matches the proportion of extinct species that were passerines,  also indicate that these extinct species were more functionally similar than expected, thus relatively \textit{increasing} trait diversity (i.e.
relative to random extinctions).
In addition, it is worth stressing that these analyses included marine species (which constitutes less than 4\% of  known extinctions in Hawaii, while making up 11\% of the prehistoric assemblage) and different results may have been found if focusing exclusively on terrestrial birds.

Following our results and these thoughts, we would like to encourage workers in ecology and evolution to test whether their metrics are capable of capturing a pattern of interest before discussing the process and/or mechanism inferred by it.
We encourage workers to test their metrics using simplified example datasets to see if it has the ability to capture at least crude changes as exemplified in this work (e.g. using simulations \citealt{guillerme2024treats} or the \texttt{moms} interactive package) \citealt{guillerme2020shifting}.
Importantly, it should always be borne in mind that empirical mechanisms ought of course to be more complex in real world scenarios.
For example, evolutionary mechanisms can vary through time or per clades (and so patterns are often the results of multiple processes); or ecological mechanisms are often intertwined and work together to generate a pattern (e.g. facilitation + competition; \citealt{danet2024species}), or counteract each others by operating on the trait space in opposite directions (e.g., competition + filtering; \citealt{mammola2024functional})

\section{Conclusion}
Different ecological and evolutionary processes or mechanisms do not always lead to different patterns.
Different observed patterns can originate from the same processes and mechanisms.
Our results based on simulations and empirical data suggest that the same data can be interpreted differently depending on the choice of trait space occupancy metric (disparity, dissimilarity or functional diversity metrics).
Different metrics are designed to capture different aspects \citep{guillerme2020shifting,mammola2021concepts} but also perform better or worse at their designed task depending on the mechanism or process of interest.
Using space occupancy metrics for describing functional diversity is tricky. This is because the metrics have properties that can be counterintuitive based on the data at hand. And also because functions of ecosystems or organisms and niches within it are hard to capture/understand/define. This means that measuring functional diversity is based on both the definition of the metric and the function of interest and thus cannot be served by a one-fits-all metric.
In other words, it is important to choose a disparity/dissimilarity/functional diversity metric depending on the question and the data at hand rather than as a default option or solely based on a previous inspiring publication.

@@

\section{Acknowledgements}
Thanks to Andrew Beckerman, Natalie Cooper, Alain Danet, Thomas Johnson and Gavin Thomas for comments on early versions of this manuscripts.\\
TG was funded by the UKRI-NERC grant NE/X016781/1.
